---
title: "Calling Translocations"
author: 
  - name: "Teun van den Brand"
    email: "t.vd.brand@nki.nl"
    affiliation: "Netherlands Cancer Institute"
date: '`r format(Sys.time(), "%d/%m/%Y")`'
output: 
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
params:
  donor: P11_LM
  resolution: 500000
  threshold: 100
bibliography: /DATA/users/t.vd.brand/bibliography/articles.bibtex
knit: |
  # Redirect output to HTML folder
  (function(inputFile, encoding) {
    outputFile <- gsub("Rmd$", "html", basename(inputFile))
    rmarkdown::render(
      input = inputFile,
      encoding = encoding,
      output_file = here::here("html", outputFile)
    )
  })
---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Load packages

```{r stfu, include = FALSE}
# No, I do *not* want your start-up messages, thank you very much
library <- function(...) {
  suppressPackageStartupMessages(base::library(...))
}
```


```{r load_packages}
library(here)
library(ggplot2)
library(scales)
library(GENOVA)
library(data.table)
library(EBImage)

source(here("rscripts", "project_functions.R"))
```

### Load data

```{r load_data}
sample <- donor <- params$donor
files <- list.files(
  here("data", "hic", "pools"),
  full.names = TRUE, pattern = donor
)
exp <- list(load_contacts(
  signal_path = files[1],
  resolution = params$resolution,
  scale_bp = NULL,
  balancing = FALSE
))
names(exp) <- donor
exp[[donor]]$MAT <- exp[[donor]]$MAT[V3 != 0]
```

### Setup Aesthetics

```{r setup_aes}
mycolour <- "#000000FF" # Opaque Black

theme_set(theme_gray())
theme_update(text = element_text(colour = mycolour),
             line = element_line(colour = mycolour),
             aspect.ratio = 1,
             axis.line  = element_line(colour = mycolour),
             axis.ticks = element_line(colour = mycolour),
             axis.text  = element_text(colour = mycolour),
             legend.key = element_blank(),
             legend.background = element_rect(colour = NA, fill = NA),
             panel.background = element_blank(),
             panel.grid.major = element_line(colour = "grey95"),
             panel.grid.minor = element_blank(),
             plot.background  = element_blank(),
             strip.background = element_blank(),
             strip.text = element_text(colour = mycolour))
CYRUP <- c("#fed500", "#009bef", "#ff5c49", "#949494", "#D8D8D8")
rm(mycolour)
```

### Functions

```{r}
sweep_chroms <- function(data, chroms, bins, FUN, ...) {
  
  grid <- t(combn(chroms, 2))
  chr_idx <- split(bins$index, bins$V1)
  
  for (i in seq_len(nrow(grid))) {
    
    chr1 <- chr_idx[[grid[i, 1]]]
    chr2 <- chr_idx[[grid[i, 2]]]
    
    dat <- data[chr1, chr2]
    
    res <- FUN(dat, ...)
    
    data[chr1, chr2] <- res
  }
  
  data
}
```

## Aim

Try to detect translocations in patient Hi-Cs.

## Raw

```{r}
good_chroms <- paste0("chr", c(1:22, "X"))
bins <- exp[[donor]]$IDX
bins <- bins[bins$V1 %in% good_chroms]
bins <- bins[order(factor(bins$V1, good_chroms))]

data <- exp[[donor]]$MAT

# Filter out under-utilised bins
obs_per_bin <- integer(nrow(bins))
t1 <- table(data$V1)
t1 <- t1[as.integer(names(t1)) %in% bins$V4]
obs_per_bin[match(as.integer(names(t1)), bins$V4)] <- as.vector(t1)
t2 <- table(data$V2)
t2 <- t2[as.integer(names(t2)) %in% bins$V4]
obs_per_bin[match(as.integer(names(t2)), bins$V4)] <- 
 obs_per_bin[match(as.integer(names(t2)), bins$V4)] + as.vector(t2)

bins <- bins[obs_per_bin > 2000,]
bins[, index := seq_len(nrow(bins))]

# Set trans-join
join <- CJ(bins$V4, bins$V4)
join <- join[V2 >= V1]
join <- join[!bins$V1[match(V1, bins$V4)] == bins$V1[match(V2, bins$V4)]]

# Select data
data <- data[join]
data[is.na(V3), V3 := 0]

data[, V1 := match(V1, bins$V4)]
data[, V2 := match(V2, bins$V4)]
```

### Display

```{r}
df <- rbind(
  data,
  setNames(data, c("V2", "V1", "V3"))[, c("V1", "V2", "V3"), with = FALSE]
)

ggplot(df, aes(V1, V2, fill = V3)) +
  geom_raster() +
  scale_fill_cis(
    name = "Contacts",
    limits = c(0, 50)
  ) +
  scale_y_continuous(
    trans = "reverse"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  ggtitle("Raw contacts") +
  theme_void() +
  theme(aspect.ratio = 1)
```

## De-CNV

```{r}
tab <- table(c(data$V1, data$V2))
t1 <- data[, .(V3 = sum(V3)), by = "V1"]
t2 <- data[, .(V3 = sum(V3)), by = "V2"]
bin_means <- merge.data.table(t1, t2, by.x = "V1", by.y = "V2", all = TRUE)
bin_means[is.na(V3.x), V3.x := 0]
bin_means[is.na(V3.y), V3.y := 0]
bin_means <- bin_means[, .(V1, V3 = V3.x + V3.y)]
bin_means[, V3 := V3 / as.vector(tab)[V1]]
bin_means[, V3 := V3 / sum(V3)]

total_sum <- sum(data$V3)

data[, bin_x := log10(bin_means$V3)[match(V1, bin_means$V1)]]
data[, bin_y := log10(bin_means$V3)[match(V2, bin_means$V1)]]

set.seed(42)
sample <- sample(nrow(data), 1e5)
fit <- MASS::glm.nb(
  V3 ~ bin_x * bin_y,
  data = data, subset = sample
)

pseudo_count <- 20
pred <- predict(fit, newdata = data, type = 'response')
data$res <- log2((data$V3 + pseudo_count) / (pred + pseudo_count))
```

### Display

```{r}
df <- data[, c("V1", "V2", "res"), with = FALSE]

df <- rbind(
  df,
  setNames(df, c("V2", "V1", "res"))[, c("V1", "V2","res"), with = FALSE]
)

ggplot(df, aes(V1, V2, fill = res)) +
  geom_raster() +
  scale_fill_gradientn(
    colours = GENOVA:::bezier_corrected_divergent,
    name = "Copy number\nCorrected",
    limits = c(-2, 2), oob = oob_squish
  ) +
  scale_y_continuous(
    trans = "reverse"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  ggtitle("Corrected contacts") +
  theme_void() +
  theme(aspect.ratio = 1)
```

## Decompartmentalise

```{r}
mat <- matrix(0, nrow(bins), nrow(bins))
mat[data[, cbind(V1, V2)]] <- data$res
mat[data[, cbind(V2, V1)]] <- data$res

eig <- RSpectra::eigs_sym(mat, 1, which = "LA")
eig <- tcrossprod(eig$vectors)

data[, eig := eig[cbind(V1, V2)]]
fit <- MASS::rlm(res ~ 0 + eig, data = data)
data[, eigres := residuals(fit)]

df <- data[, c("V1", "V2", "eigres"), with = FALSE]
df <- rbind(
  df,
  setNames(df, c("V2", "V1", "eigres"))
)

ggplot(df, aes(V1, V2, fill = eigres)) +
  geom_raster() +
  scale_fill_gradientn(
    colours = GENOVA:::bezier_corrected_divergent,
    name = "Contacts",
    limits = c(-6, 6), oob = oob_squish
  ) +
  scale_y_continuous(
    trans = "reverse"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  ggtitle("Decompartmentalised") +
  theme_void() +
  theme(aspect.ratio = 1)
```

## Use kernels

### Define kernels

```{r}
kern1 <- matrix(c(
  75,  150, 300,  0, -300,  -150, -75,
  150, 300, 600,  0, -600,  -300, -150,
  300, 600, 1200, 0, -1200, -600, -300,
  0,   0,   0,    0, 0,     0,    0,
  -300, -600, -1200, 0, -1200, -600, -300,
  -150, -300, -600,  0, -600,  -300, -150,
  -75,  -150, -300,  0, -300, -150, -75
), 7, 7)

kern2 <- apply(kern1, 1, rev)
kern3 <- t(kern2)
kern4 <- kern1
kern4[] <- rev(kern4)

kernels <- list(kern1, kern2, kern3, kern4)

df <- lapply(kernels, reshape2::melt)
df <- rbindlist(df, idcol = "kernel")

ggplot(df, aes(Var1, Var2, fill = value)) +
  geom_raster() +
  geom_text(aes(label = value), size = 3) +
  scale_fill_GENOVA_div(
    midpoint = 0, name = "Weight"
  ) +
  facet_wrap(~ kernel) +
  coord_cartesian(expand = FALSE) +
  labs(x = "x", y = "y") +
  theme(panel.border = element_rect(colour = "black", fill = NA))
```

### Sweep

```{r}
mat[] <- 0
mat[data[, cbind(V1, V2)]] <- pmax(data$eigres, 0)
lim <- range(mat)

new_mat <- sweep_chroms(mat, good_chroms, bins, function(img) {
  img <- Image(rescale(img, from = lim), dim = dim(img), "Grayscale")
  img <- medianFilter(img, 3)
  img <- lapply(kernels, function(k) {
    pmax(imageData(filter2(img, k, "replicate")), 1)
  })
  Reduce(`+`, img)
})

```

### Display

```{r}
df <- reshape2::melt(new_mat)

ggplot(df, aes(Var1, Var2, fill = value)) +
  geom_raster() +
  scale_fill_viridis_c(limits = c(0, 1000), oob = oob_squish) +
  scale_y_reverse() +
  theme_void() +
  theme(aspect.ratio = 1)
```

## Thresholding

```{r}
new_upper <- data.table(
  x = row(new_mat)[upper.tri(new_mat)],
  y = col(new_mat)[upper.tri(new_mat)],
  z = new_mat[upper.tri(new_mat)]
)

vals <- sort(as.vector(new_upper$z))
diff <- diff(vals)

rle <- rle(diff > mean(diff))
rle$values[rle$lengths == 1] <- !rle$values[rle$lengths == 1]
rle <- rle(rep.int(rle$values, rle$lengths))
ends <- cumsum(rle$lengths)
starts <- ends - rle$lengths + 1
thres <- vals[tail(starts, 1)]

test_thres <- seq(0.5, 3, by = 0.25)
```

### Exploring thresholds

```{r}
img <- Image(new_mat, dim = dim(new_mat))
thres_stats <- lapply(test_thres, function(thr) {
  rel <- thr
  thr <- rel * thres
  
  img <- sweep_chroms(
    img > thr, good_chroms, bins,
    function(img) {
      dilate(Image(img), makeBrush(21, "box"))
    }
  )
  
  labels <- bwlabel(img)
  tab <- table(labels)
  data.table(
    id = as.numeric(names(tab)),
    n_pixels = as.vector(tab),
    abs_thres = thr,
    rel_thres = rel
  )
})

thres_stats <- rbindlist(thres_stats)
thres_stats <- thres_stats[id != 0]
smmry <- thres_stats[, .(n = max(id), avg_pixels = mean(n_pixels)), by = "rel_thres"]

ggplot(smmry, aes(rel_thres, n)) +
  geom_point(aes(size = avg_pixels)) +
  # geom_vline(xintercept = params$threshold, linetype = "dashed") +
  # geom_hline(yintercept = 27, linetype = "dashed") +
  scale_y_continuous(
    trans = "log10"
  ) +
  scale_x_continuous(
    sec.axis = sec_axis(
      ~ .x * thres, name = "Absolute threshold"
    )
  ) +
  labs(
    x = "Relative Threshold",
    y = "Number of identified translocations"
  )
```

### Apply chosen threshold

```{r}
labels <- dilate(img > params$threshold, makeBrush(21, "box"))
labels <- bwlabel(labels)

ind <- which(labels > 0)
ind <- arrayInd(ind, dim(labels))
labels <- labels[ind]
labels <- data.table(
  V1 = pmin(ind[, 1], ind[, 2]),
  V2 = pmax(ind[, 1], ind[, 2]),
  group = labels
)
```

### Translate to pixel

```{r}
raw <- data[labels, on = c("V1", "V2")]
raw <- raw[, .SD[which.max(V3)], by = "group"]
raw <- raw[, c("V1", "V2", "V3"), with = FALSE]
raw <- raw[V3 > 10 * mean(data$V3) & V3 > 20]
xmatch <- match(raw$V1, bins$index)
raw[, chr_x   := bins$V1[xmatch]]
raw[, start_x := bins$V2[xmatch]]
raw[, end_x   := bins$V3[xmatch]]
raw[, bin_x   := bins$V4[xmatch]]
ymatch <- match(raw$V2, bins$index)
raw[, chr_y   := bins$V1[ymatch]]
raw[, start_y := bins$V2[ymatch]]
raw[, end_y   := bins$V3[ymatch]]
raw[, bin_y   := bins$V4[ymatch]]

fwrite(
  raw,
  file = here::here(
    "processed_data", "translocations",
    paste0(donor, "_computed.tsv")
  ),
  sep = "\t"
)
```

### Display

```{r}
df <- data[, c("V1", "V2", "V3"), with = FALSE]
df <- rbind(
  df,
  setNames(df, c("V2", "V1", "V3"))[, c("V1", "V2", "V3"), with = FALSE]
)

ggplot(df, aes(V2, V1, fill = V3)) +
  geom_raster() +
  geom_point(
    data = raw,
    shape = 1,
    colour = "dodgerblue",
    stroke = 1, size = 3
  ) +
  scale_fill_cis(
    name = "Contacts",
    limits = c(0, 50)
  ) +
  scale_y_continuous(
    trans = "reverse"
  ) +
  labs(
    x = NULL,
    y = NULL
  ) +
  ggtitle("Raw contacts") +
  theme_void() +
  theme(aspect.ratio = 1)
```

## References

<div id="refs"></div>

## Session Info

```{r session_info, include=TRUE, echo=FALSE, results = "asis"}
pkgs <- loadedNamespaces()
libTvdB::format_sessioninfo_html(pkgs)
```
