---
title: "Untitled"
author: 
  - name: "Teun van den Brand"
    email: "t.vd.brand@nki.nl"
    affiliation: "Netherlands Cancer Institute"
date: '`r format(Sys.time(), "%d/%m/%Y")`'
output: 
  html_document:
    theme: united
    highlight: pygments
    fig_caption: yes
    code_folding: hide
    df_print: kable
    toc: true
    toc_depth: 4
    toc_float:
      collapsed: false
bibliography: /DATA/users/t.vd.brand/bibliography/articles.bibtex
knit: |
  # Redirect output to HTML folder
  (function(inputFile, encoding) {
    outputFile <- gsub("Rmd$", "html", basename(inputFile))
    rmarkdown::render(
      input = inputFile,
      encoding = encoding,
      output_file = here::here("html", outputFile)
    )
  })
---

## Setup {.tabset}

<!-- little HTML script to do indentation of the table of contents -->
<script>
    $(document).ready(function() {
      $items = $('div#TOC li');
      $items.each(function(idx) {
        num_ul = $(this).parentsUntil('#TOC').length;
        $(this).css({'text-indent': num_ul * 10, 'padding-left': 0});
      });

    });
</script>

```{css, echo = FALSE}
div.sourceCode {
  overflow-x: hidden;
}
```

### Load packages

```{r stfu, include = FALSE}
# No, I do *not* want your start-up messages, thank you very much
library <- function(...) {
  suppressPackageStartupMessages(base::library(...))
}
```


```{r load_packages}
library(ggplot2)
library(scales)
library(here)
library(data.table)
library(GenomicRanges)
library(patchwork)
library(Matrix)
source(here("rscripts", "project_functions.R"))
```

### Load data

```{r load_data}
tads <- readRDS(here("rds", "01_TADs.rds"))
tads$exp <- sample_lut[tads$exp]
tads <- tads[grepl("_PE$", tads$exp),,]

cistrome_dir <- "/DATA/users/moreno/references/cistrome_human"
```

### Setup Aesthetics

```{r setup_aes}
set_figure_theme()
```

## Aim

```{r}
borders <- GRanges(
  seqnames = c(tads$chrom, tads$chrom),
  IRanges(c(tads$start, tads$end), width = 1),
  exp = c(tads$exp, tads$exp)
)

wide_borders <- resize(borders, 3 * 20e3, fix = "center")
wide_borders <- reduce(wide_borders)
wide_borders <- wide_borders[width(wide_borders) <= 1e5]

binary <- vapply(
  split(borders, borders$exp),
  function(gr) {
    overlapsAny(wide_borders, gr)
  },
  logical(length(wide_borders))
)
rownames(binary) <- paste0("TAD_", seq_len(nrow(binary)))
wide_borders$freq <- rowSums(binary)
```

```{r}
tab <- table(wide_borders$freq)
print(sum(tab))
plot(tab)
```

### Metadata import


```{r}
cistrome <- list.files(cistrome_dir, full.names = TRUE, pattern = "QC.txt")
cistrome <- setNames(lapply(cistrome, fread), c("Factor", "Histones"))
cistrome <- rbindlist(cistrome, idcol = "Data_type")
```

### Inclusion criteria

```{r}
cell_line <- c("MCF7", "MCF-10A", "MDA-MB-231", "T47D", "T47D-MTVL")
cell_type <- c(
  "Epithelium", "Mammary carcinoma cell", "Human Breast Cancer Cell"
)
tissue    <- c("Breast", "Mammary Gland")

cistrome[, keep := FALSE]
cistrome[Cell_type %in% cell_type, keep := TRUE]
cistrome[Cell_line %in% cell_line, keep := TRUE]
cistrome[Tissue_type %in% tissue,  keep := TRUE]

meta <- cistrome[keep == TRUE]
```

### Exclusion criteria

```{r}
# Anything that is not
tissue <- c("Breast", "None", "Mammary Gland")
meta <- meta[Tissue_type %in% tissue]

# For each cell_line/factor combination pick dataset with most strong peaks
meta <- meta[, .SD[which.max(PeaksFoldChangeAbove10)], 
             by = c("Cell_line", "Factor")]
```

### Overview

#### Tissue type

```{r}
phi   <- 2/(1 + sqrt(5))
width <- phi / 2
pos   <- position_dodge(phi, preserve = "single")
fmt   <- function(x, n = 10) {
  forcats::fct_rev(forcats::fct_lump(forcats::fct_infreq(x), n))
}

ggplot(meta, aes(y = fmt(Tissue_type), fill = Data_type)) +
  geom_bar(width = width, position = pos) +
  scale_x_continuous(expand = c(0, 0, 0.1, 0)) +
  labs(
    x = "# Datasets",
    y = "Tissue type",
    fill = "Dataset type"
  ) +
  theme(
    panel.grid.major.y = element_blank()
  )
```

#### Cell type

```{r}
ggplot(meta, aes(y = fmt(Cell_type), fill = Data_type)) +
  geom_bar(width = width, position = pos) +
  scale_x_continuous(expand = c(0, 0, 0.1, 0)) +
  labs(
    x = "# Datasets",
    y = "Cell type"
  ) +
  theme(
    panel.grid.major.y = element_blank()
  )
```

#### Cell line

```{r}
ggplot(meta, aes(y = fmt(Cell_line), fill = Data_type)) +
  geom_bar(width = width, position = pos) +
  scale_x_continuous(expand = c(0, 0, 0.1, 0)) +
  labs(
    x = "# Datasets",
    y = "Cell line"
  ) +
  theme(
    panel.grid.major.y = element_blank()
  )
```

#### Factors 

```{r}
ggplot(meta, aes(y = fmt(Factor), fill = Data_type)) +
  geom_bar(width = width, position = pos) +
  scale_x_continuous(expand = c(0, 0, 0.1, 0)) +
  labs(
    x = "# Datasets",
    y = "Cell type"
  ) +
  theme(
    panel.grid.major.y = element_blank()
  )
```

## Import Cistrome

### Match files

```{r}
cistrome_files <- list.files(
  cistrome_dir, recursive = TRUE, full.names = TRUE
)
dcid <- basename(cistrome_files)
dcid <- tstrsplit(dcid, "_", keep = 1)[[1]]
meta_match <- match(as.character(meta$DCid), dcid)
meta$file <- cistrome_files[meta_match[!is.na(meta_match)]]

# Kick out empty files
meta <- meta[fs::file_size(file) > 100]
meta[, keep := NULL]
```

```{r}
cistrome_data <- lapply(
  setNames(meta$file, meta$DCid),
  fread, sep = "\t", header = FALSE
)
cistrome_data <- rbindlist(cistrome_data, idcol = "dcid", fill = TRUE)
```

### Kick out small datasets

```{r}
good_chroms <- paste0("chr", c(1:22, "X"))
cistrome_data <- cistrome_data[V1 %in% good_chroms, ]

npeaks <- cistrome_data[, .(n = .N), by = "dcid"]
keep <- npeaks[n >= 1000, dcid]

meta <- meta[DCid %in% keep]
cistrome_data <- cistrome_data[dcid %in% keep]
cistrome_data <- cistrome_data[, 
  GRanges(V1, IRanges(V2 + 1, V3), DCid = dcid, seqinfo = seqinfo(wide_borders))
]
cistrome_data <- sort(cistrome_data)
```

### Count at TAD borders

```{r}
overlaps <- findOverlaps(cistrome_data, wide_borders)
overlaps <- data.table(
  peak   = from(overlaps),
  border = to(overlaps)
)
overlaps[, dataset := cistrome_data$DCid[peak]]
overlaps <- overlaps[, .N, by = c("dataset", "border")]
overlaps[, id := match(dataset, meta$DCid)]

counts <- overlaps[,
  sparseMatrix(
    i = border, j = id, x = 1, 
    dims = c(length(wide_borders), nrow(meta))
  )
]

colnames(counts) <- meta$DCid
counts <- as.matrix(counts)
```


```{r}
freq <- rowsum(counts, wide_borders$freq)
chisq <- chisq.test(freq)
obsexp <- (chisq$observed + 1) / (chisq$expected + 1)

cmax <- matrixStats::colMaxs(obsexp)
cmin <- matrixStats::colMins(obsexp)

extreme10 <- unique(c(
  which(order(order(-cmax)) <= 10),
  which(order(order(cmin)) <= 10)
))

df <- obsexp[, extreme10]
clust <- hclust(dist(t(df)))
ord <- colnames(df)[clust$order]

df <- reshape::melt(df)

i <- match(df$X2, meta$DCid)
df <- cbind(df, meta[i,])

ggplot(df, aes(factor(X1), factor(DCid), fill = value)) +
  geom_raster() +
  scale_fill_gradientn(
    colours = div_colours,
    rescale = ~scales::rescale_mid(.x, mid = 1)
  ) +
  scale_x_discrete(name = "Frequency", expand = c(0, 0)) +
  scale_y_discrete(
    name = NULL, expand = c(0, 0),
    limits = ord,
    labels = function(i) {
      i <- match(i, meta$DCid)
      paste0(meta$Factor[i], "\n", meta$Cell_line[i])
    }
  ) +
  coord_equal()
```

## model


```{r}
res <- lapply(seq_len(ncol(obsexp)), function(i) {
  df <- data.frame(x = seq_len(nrow(obsexp)), y = obsexp[,i])
  fit0 <- lm(y ~ 1, data = df)
  fit1 <- lm(y ~ x, data = df)
  coef <- coef(fit1)[2]
  ll_stat <- -2 * (as.numeric(logLik(fit0)) - as.numeric(logLik(fit1)))
  pval <- pchisq(ll_stat, df = 1, lower.tail = FALSE)
  data.frame(i = i, coef = coef, pval = pval)
})
res <- rbindlist(res)

ggplot(res, aes(coef, -log10(p.adjust(pval, "fdr")))) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), linetype = 2)
```

```{r}
extreme10 <- unique(c(
  which(order(order(-res$coef)) <= 10),
  which(order(order(res$coef)) <= 10)
))

df <- obsexp[, extreme10]
clust <- hclust(dist(t(df)))
ord <- colnames(df)[clust$order]

df <- reshape::melt(df)

i <- match(df$X2, meta$DCid)
df <- cbind(df, meta[i,])

ggplot(df, aes(factor(X1), factor(DCid), fill = value)) +
  geom_raster() +
  scale_fill_gradientn(
    name = "Obs + 1 /\nExp + 1",
    colours = div_colours,
    rescale = ~scales::rescale_mid(.x, mid = 1)
  ) +
  scale_x_discrete(name = "Frequency", expand = c(0, 0)) +
  scale_y_discrete(
    name = NULL, expand = c(0, 0),
    limits = ord,
    labels = function(i) {
      i <- match(i, meta$DCid)
      paste0(meta$Factor[i], "\n", meta$Cell_line[i])
    }
  ) +
  coord_equal()
```

## Save stuff

```{r}
wide_borders <- trim(wide_borders)
wide_borders <- as.data.frame(wide_borders)

fwrite(
  wide_borders, file = here("references", "TAD_borders.bed"),
  sep = "\t", row.names = FALSE, col.names = FALSE
)


tab <- cbind(meta, res)
tab$fdr <- p.adjust(tab$pval, "fdr")
fwrite(tab, file = here("references", "TAD_border_cistrome_stats.tsv"), sep = "\t")
```


## References

<div id="refs"></div>

## Session Info

```{r session_info, include=TRUE, echo=FALSE, results = "asis"}
# pkgs <- loadedNamespaces()
# libTvdB::format_sessioninfo_html(pkgs)
```
